---
title: "Steps taken to generate Figure 2 of Koeber et al, Nature 2026"
output: html_document
date: "2024-04-02"
---


# This document details the methods/steps used to generate figures used in Figure 2 (Koeber et al. Nature, 2026):
## Load required libraries
```{r}
.p <- c("tidyverse", "ggpubr", "googlesheets4", "data.table", "regioneR", "VennDiagram")
lapply(.p, require, character.only =T)

```

## Figure A is created outside of R, using Adobe Illustrator + BioRender

## Figure B: Bar chart comparing the proportion of co-bound SOX2 and SOX9 peaks in each GSC culture
```{r}
Sox_peaks = fread("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/basic_overlap_stats.txt")

b <- Sox_peaks %>% filter(Sample != "Random") %>%
  select(Sample, Sox2 = `%_Sox2_overlap`, Sox9= `%_Sox9_overlap`) %>%
  mutate(Sox2 = Sox2*100, Sox9 = Sox9*100) %>%
  mutate(Sample = as.factor(Sample)) %>%
  # mutate(Sample = fct_relevel(Sample, "E27","E21","E34","E31","E28","E17","E37")) %>%
  pivot_longer(!Sample) %>%
  ggplot() +
  geom_bar(aes(x=Sample, y=value, fill= name), position = "dodge", stat = "identity") +
  labs(x=NULL,y = "Proportion within\n co-bound peaks (%)", fill = "Transcription factor") +
  scale_fill_manual(values = c("Sox2" = "grey", "Sox9" = "red")) +
  scale_y_continuous(breaks = c(0,50,100)) + 
  theme_classic() +
  theme(panel.grid = element_blank(), 
        legend.position = "top",
        legend.text = element_text(size=12), 
        legend.title = element_text(size=13, face = "bold"), 
        axis.text.y = element_text(size=12),
        axis.text.x = element_text(size=11),
        axis.title.y= element_text(size=14))

pdf("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/figures/proportion_co_bound_peaks.pdf", height = 2.3, width = 3.5)
b
dev.off()
```

## Figure C: Venn diagram of consensus SOX2 and SOX9 overlap:
```{r}
## Load consensus sets:
### Co-bound set:
co_bound_consensus <- read.delim("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/Merged_rep/Macs2/Consensus_bed_files/Merged_rep_Consensus5_overlap.bed", header = F) %>%
  toGRanges() %>%
  reduce()
length(co_bound_consensus)

## SOX2 exclusive:
Sox2_consensus <- read.delim("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/Merged_rep/Consensus_bed_files/Merged_rep_consensus5_Sox2_exclusive.bed", header=F) %>%
  toGRanges() %>%
  reduce()
length(Sox2_consensus)

## SOX9 exclusive:
Sox9_consensus <- read.delim("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/Merged_rep/Consensus_bed_files/Merged_rep_consensus5_Sox9_exclusive.bed", header=F) %>%
  toGRanges() %>%
  reduce()
length(Sox9_consensus)


pdf("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/figures/venn_diag_co_bound.pdf", height = 3, width = 3)
draw.pairwise.venn(area1 = length(Sox2_consensus) + length(co_bound_consensus), 
                   area2 = length(Sox9_consensus) + length(co_bound_consensus), 
                   cross.area = length(co_bound_consensus), scaled =T, 
                   fill = c("darkgrey","brown"), 
                   cex = c(1.5,1.5,1.5), 
                   alpha = c(0.5,0.5), fontfamily = rep("sans",3))

dev.off()

## Circular permutation test:
### Get all SOX2 peaks:
sox2_peaks <- c(Sox2_consensus, co_bound_consensus)

### Create all SOX9 peaks:
sox9_peaks <- c(Sox9_consensus, co_bound_consensus)

### Perform overlap test:
perm_test_co_bound <- permTest(A = sox2_peaks, 
                               B = sox9_peaks,
                               ntimes = 1000,
                               randomize.function=circularRandomizeRegions, 
                               evaluate.function=numOverlaps, 
                               count.once=TRUE, 
                               genome="hg38", 
                               mc.set.seed=FALSE, 
                               mc.cores=16)
perm_test_co_bound
```

## Figure D: Venn diagram of consensus co-bound sites vs super-enhancers in GSC:
```{r}
## Load consensus SE set: SEs occuring in >1 GSC
consensus_SE <- read.table("Data/ChIP-seq/H3K27ac/SuperEnhancers/Consensus_SEs.bed", header = F) %>% 
  toGRanges() %>%
  reduce()
length(consensus_SE)

## Load consensus co-bound SOX2 and SOX9 set:
co_bound_consensus <- read.delim("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/Merged_rep/Consensus_bed_files/Merged_rep_Consensus5_overlap.bed", header = F) %>%
  toGRanges() %>%
  reduce()
length(co_bound_consensus)

co_bound_SOX_SEs <- subsetByOverlaps(co_bound_consensus, consensus_SE)
co_bound_SOX_not_SE <- subsetByOverlaps(co_bound_consensus, consensus_SE, invert = T)
SE_not_co_bound <- subsetByOverlaps(consensus_SE, co_bound_consensus, invert = T)

pdf("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/figures/venn_diag_co_bound_SEs.pdf", height = 3, width = 3)
draw.pairwise.venn(area2 = length(co_bound_SOX_not_SE) +length(co_bound_SOX_SEs), 
                   area1 = length(SE_not_co_bound) + length(co_bound_SOX_SEs), 
                   cross.area = length(co_bound_SOX_SEs), scaled =T, 
                   cex = c(1.5,1.5,1.5), 
                   fill = c("lightblue","brown"), alpha = c(0.5,0.5), inverted = T, fontfamily = rep("sans",3))


dev.off()

### Perform overlap test:
perm_test_co_bound_SE <- permTest(A = co_bound_consensus, 
                               B = consensus_SE,
                               ntimes = 1000,
                               randomize.function=circularRandomizeRegions, 
                               evaluate.function=numOverlaps, 
                               count.once=TRUE, 
                               genome="hg38", 
                               mc.set.seed=FALSE, 
                               mc.cores=16)


```

## Figure E: Signal intensity at co-bound sites
```{r}
files <- list.files("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/Signal_intensity/", pattern = "_scores_per_bin.tab", full.names = F)
files <- files[grep(pattern = "Sox",  files, invert = T)]

signal_intensity <- lapply(seq_along(files), function(x) {
  name <- str_split(gsub(".*/", "", files[x]), "_")[[1]][1]
  sig <- fread(paste0("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/Signal_intensity/", files[x]), header = F, skip = 1) %>% mutate(cell_line = name) %>% filter(!str_detect(V1, "HLA-|random|Y|alt")) %>% 
    mutate(genomic_location = paste0(V1,":",V2,"-",V3))
  return(sig)
  
}) %>% rbindlist()

N <- 14
e <- signal_intensity %>% 
  dplyr::rename(Sox2 = V4, Sox9 = V5) %>% 
   mutate(Sox2 = if_else(Sox2 <0, 0, Sox2),
          Sox9 = if_else(Sox9 <0, 0, Sox9)) %>% 
  ggplot(aes(x=Sox2, y = Sox9)) + 
  geom_bin2d(bins=50) +
  scale_fill_viridis_c(option = "magma")+ 
  labs(x="\nSox2 signal intensity", y="Sox9 signal intensity\n", fill="Count") +
  facet_grid(~cell_line, scales = "free") + 
  stat_cor(method = "pearson", aes(label = ..r.label..), label.y.npc = 0.95, label.x.npc = 0.05, size=5) +
  # stat_cor(method = "pearson", label.y.npc = 0.95, label.x.npc = 0.05, size=5) +
  scale_y_log10() + 
  scale_x_log10() + 
  theme_minimal() +
  theme(strip.text = element_text(size=N+2),
        strip.background = element_blank(),
        legend.text = element_text(size=N), 
        panel.grid = element_blank(), 
        legend.title = element_text(size=N+1, face = "bold"),
        axis.text = element_blank(), 
        axis.ticks = element_blank(),
        axis.title = element_text(size=N+2))

pdf("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/figures/signal_intensity_sox2_sox9.pdf", height = 3, width = 10, useDingbats = F, pointsize = 1)
e
dev.off()


```

## Figure F: GREAT GO terms:
```{r}
## Biological context:
go_biological <- read_tsv("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/Merged_rep/Gene_set_Ontologies/GREAT_analysis/GREAT_GO_BIOLOGICAL.tsv", skip = 3)
go_molecular <- read_tsv("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/Merged_rep/Gene_set_Ontologies/GREAT_analysis/GREAT_GO_MOLECULAR.tsv", skip = 3)
go_cellular <- read_tsv("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/Merged_rep/Gene_set_Ontologies/GREAT_analysis/GREAT_GO_CELLULAR.tsv", skip = 3)

colnames(go_biological)
x <- go_biological %>% 
  dplyr::select(Desc, BinomFdrQ, Genes) %>% 
  arrange(BinomFdrQ) %>% 
  slice_head(n=5) %>% 
  mutate(type = "GO Biological")

y <- go_molecular %>% 
  dplyr::select(Desc, BinomFdrQ, Genes) %>% 
  arrange(BinomFdrQ) %>% 
  slice_head(n=5) %>% 
  mutate(type = "GO Molecular")

z <- go_cellular %>% 
  dplyr::select(Desc, BinomFdrQ, Genes) %>% 
  arrange(BinomFdrQ) %>% 
  slice_head(n=5) %>% 
  mutate(type = "GO Cellular")

f <- rbind(x,y,z) %>% 
  ggplot(aes(x=-log10(BinomFdrQ), y= reorder(Desc, -BinomFdrQ))) + 
  geom_col(fill = "darkgrey") +
  theme_classic() +
  lemon::coord_capped_cart(left = "both", bottom = "both") +
  facet_wrap(~type, ncol=1, scales = "free_y") +
  scale_x_continuous(breaks= c(0,2,4), limits=c(0,5), expand = c(0,0.1)) + 
  scale_y_discrete(expand=c(0.2,0)) +
  labs(x=expression("-log"[10]*"Q-value"), y = NULL) + 
  theme(strip.text = element_text(hjust = 0, face = "bold", size=11),
        strip.background = element_blank(),
        axis.text.y = element_text(size=11, face = "italic"),
        axis.text.x = element_text(size=11),
        panel.grid = element_blank())

pdf("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/figures/GREAT_GO_terms.pdf", height = 4, width = 5.3)
f
dev.off()


# 
# ## Get all genes assigned to co-bound peaks:
# genes_co_bound <- read_tsv("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/Merged_rep/Gene_set_Ontologies/GREAT_analysis/GREAT_genes.txt", col_names = F, skip = 1) %>% 
#   separate(X2, into = paste("A", 1:20, sep=""), sep =",*() ") %>% 
#   pivot_longer(!X1) %>% 
#   filter(!str_detect(value, "[(]")) %>% 
#   dplyr::select(value) %>% 
#   distinct(value) %>% 
#   filter(!str_detect(value, "ENSG")) %>% 
#   arrange(value) 
# 
# genes_co_bound %>% write_tsv("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/Merged_rep/Gene_set_Ontologies/GREAT_analysis/GREAT_genes_tidy.txt", col_names = F)

```

## Figure G: Coverage tracks of SOX2 and SOX9 near CDK6 and PTRPRZ1, USCS browser at:

## Figure H:
```{r}
TCGA <- read.delim("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/cbioportal_sox2_sox9.txt", header=T)  %>%
  select(Sample = Sample.Id,
         Cancer_type = TCGA.PanCanAtlas.Cancer.Type.Acronym,
         Sox2_expression = SOX2..mRNA.Expression..RSEM..Batch.normalized.from.Illumina.HiSeq_RNASeqV2...log2.value...1..,
         Sox9_expression = SOX9..mRNA.Expression..RSEM..Batch.normalized.from.Illumina.HiSeq_RNASeqV2...log2.value...1..)

TCGA %>% 
  ggplot(aes(x=Sox2_expression, y= Sox9_expression)) + 
  geom_point() + 
  scale_y_log10() + 
  scale_x_log10() + 
  facet_wrap(~Cancer_type, scales = "free")

## Compute correlations:
TCGA_stats <- TCGA %>% 
  group_by(Cancer_type) %>% 
  rstatix::cor_test(vars = Sox2_expression, vars2 = Sox9_expression) %>% 
  rstatix::adjust_pvalue() 

## Compute median expression of SOX2 and SOX9 for each cancer type:
# TCGA %>% 
#   group_by(Cancer_type) %>% 
#   mutate(median_SOX2 = median(Sox2_expression), 
#          median_SOX9 = median(Sox9_expression)) %>% 
#   distinct(Cancer_type, .keep_all = T) %>% 
#   dplyr::select(Cancer_type, median_SOX2, median_SOX9) %>% 
#   left_join(TCGA_stats) %>% 
#   ggplot(aes(x=log2(median_SOX2), y = log2(median_SOX9), color = cor, size = -log10(p.adj))) + 
#   geom_point() +
#   scale_color_viridis_c()

h <- TCGA_stats %>% 
  ggplot(aes(x=reorder(Cancer_type,cor), fill=-log10(p.adj), y = cor)) +
  geom_col() + 
  scale_fill_viridis_c(option = "magma") + 
  theme_classic() +
  lemon::coord_capped_cart(left = "both") + 
  labs(
    # title = "SOX2 and SOX9 expression correlation analysis", 
       y="SOX2/SOX9\nexpression correlation", x = "\nTCGA cancer types", fill = expression("-log"[10]*"Q-value")) + 
  scale_y_continuous(breaks = c(-0.5,-0.25,0,0.25,0.5), limits = c(-0.5,0.5)) +
  theme(axis.text.y = element_text(size=12),
        axis.title = element_text(size=13),
        legend.position = "bottom",
        plot.title = element_text(size=13, face = "bold"),
        legend.title = element_text(size=13, face = "bold", vjust = 0.9),
        legend.text = element_text(size=11), 
        axis.text.x = element_text(size=12, angle = 90, hjust=0.9, vjust=0.5))
h

pdf("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/figures/TCGA_SOX2_SOX9_exp_corr.pdf", height = 4.9, width = 6)
h
dev.off()


```


