---
title: "Steps taken to generate Figure 2 of Koeber et al, Nature 2026"
author: Alhafidz Hamdan
output: html_document
date: "2026-02-01"
editor_options: 
  chunk_output_type: console
---

# This document details the methods/steps used to generate figures used in Figure 2 (Koeber et al. Nature, 2026):
## Load required libraries
```{r}
.p <- c("tidyverse", "ggpubr", "lemon", "data.table", "regioneR", "VennDiagram", "parallel", "regioneR", "rtracklayer")
lapply(.p, require, character.only =T)

theme_colplot <- theme(legend.position = "right", 
                       legend.title = element_text(size=13, face = "bold"),
                       legend.text = element_text(size=12),
                       axis.ticks.length = unit(.2, "cm"),
                       plot.margin = margin(0,0,0,0, "cm"), 
                       plot.background = element_blank(),
                       panel.background = element_blank(),
                       plot.title = element_text(size=12, face = "bold", vjust = 1),
                       axis.text = element_text(size=13), 
                       axis.title = element_text(size=14))   

```

## Figure A is created outside of R, using Adobe Illustrator + BioRender

## Figure B (left): Venn diagram of SOX2 and SOX9 overlap for a representative sample E28:
```{r}
peak_dir = "../../processed_data/"
this_sample = "E28"
this_sample_sox2 <- import(paste0(peak_dir, "/", this_sample, "_Sox2_peaks.narrowPeak")) 
this_sample_sox9 <- import(paste0(peak_dir, "/", this_sample, "_Sox9_peaks.narrowPeak")) 
 
## Plot venn diagram of overlap:
overlap <- subsetByOverlaps(this_sample_sox2, this_sample_sox9) 
sox2_exclusive <- subsetByOverlaps(this_sample_sox2, this_sample_sox9, invert = T) 
sox9_exclusive <- subsetByOverlaps(this_sample_sox9, this_sample_sox2, invert = T)  
  
pdf(paste0("Figures/", this_sample, "_sox2_sox9_overlap_venn_plot.pdf"), width = 4, height = 4, useDingbats = T, onefile = F)
draw.pairwise.venn(area1 = length(sox2_exclusive) + length(overlap), 
                   area2 = length(sox9_exclusive) + length(overlap), 
                   cross.area = length(overlap), scaled =T, 
                   fill = c("grey","darkred"), 
                   alpha = c(0.6,0.6), fontfamily = rep("sans",3))
dev.off()
  
```

## Figure B (right): SOX2 and SOX9 signal intensity heatmaps for E28, generated in bash/05_submit_deeptools_heatmap.sh
## Figure C: Bar charts of SOX2 and SOX9 peak overlaps across 7 GSC samples:
```{r}
sox2_sox9_chipseq_samples = c("E17","E21","E27","E28","E31","E34","E37")
peak_dir = "../../processed_data/macs2"
list.files(peak_dir)

## Generate sets of overlapping and exclusive peaks for each sample:
merged_peak_stats <- mclapply(seq_along(sox2_sox9_chipseq_samples), function(x) {
  
  this_sample <- sox2_sox9_chipseq_samples[x]
  this_sample_sox2 <- import(paste0(peak_dir, "/", this_sample, "_Sox2_peaks.narrowPeak")) 
  this_sample_sox9 <- import(paste0(peak_dir, "/", this_sample, "_Sox9_peaks.narrowPeak")) 
 
  overlap <- subsetByOverlaps(this_sample_sox2, this_sample_sox9) 
  overlap %>% as.data.frame() %>% 
    fwrite(paste0("../../processed_data/", sox2_sox9_chipseq_samples[x], "_Sox2_Sox9_overlap.bed"), quote=F, sep = "\t", row.names = F, col.names = F)
  sox2_only <- subsetByOverlaps(this_sample_sox2, this_sample_sox9, invert = T) 
  sox2_only %>% as.data.frame() %>% 
    fwrite(paste0("../../processed_data/", sox2_sox9_chipseq_samples[x], "_Sox2_exclusive.bed"), quote=F, sep = "\t", row.names = F)
  sox9_only <- subsetByOverlaps(this_sample_sox9, this_sample_sox2, invert = T)  
  sox9_only %>% as.data.frame() %>% 
    fwrite(paste0("../../processed_data/", sox2_sox9_chipseq_samples[x], "_Sox9_exclusive.bed"), quote=F, sep = "\t", row.names = F)
  
  dat <-   data.frame(sample = this_sample, 
                       total_sox2 = length(this_sample_sox2),
                       total_sox9 = length(this_sample_sox9),
                       sox2_only = length(sox2_only),
                       sox9_only = length(sox9_only),
                       overlap = length(overlap)) %>% 
              dplyr::mutate(perc_sox2_overlap = overlap*100/total_sox2,
                            perc_sox9_overlap = overlap*100/total_sox9)
  
  return(dat)
  
}) %>% rbindlist()

pdf("Figures/co_binding_bar_plot.pdf", width = 4.2, height = 3.1, useDingbats = T, onefile = F)
merged_peak_stats %>% 
  dplyr::select(sample, perc_sox2_overlap, perc_sox9_overlap) %>% 
  pivot_longer(!sample) %>% 
  dplyr::mutate(TF = ifelse(str_detect(name, "sox2"), "SOX2", "SOX9")) %>% 
  ggplot(aes(x=sample, y=value, fill=TF)) +
  geom_col(position = "dodge") +
  scale_y_continuous(limits=c(0,100), expand = c(0.04,0)) + 
  scale_x_discrete(expand=c(0.09,0)) +
  scale_fill_manual(values = c("SOX2" = "grey", "SOX9" = "#AD6363")) +
  labs(x="Samples", y="% of co-bound peaks", fill = "TF peaks") +
  theme_classic() + 
  lemon::coord_capped_cart(left = "both", bottom = "none") +
  theme_colplot +
  theme(legend.position = "top", plot.margin = margin(0, 0, 0, 0, "cm"))
dev.off()

```

## Figure D (left): Venn diagram of consensus SOX2 and SOX9 overlap:
```{r}
## Use bedtools multiinter to generate calls for SOX2/SOX9 overlap, SOX2 exclusive and SOX9 exclusive sets:
## For consensus overlap, use union of 5 out of 7 calls
# bedtools multiinter -header -i ChIP-seq/processed_data/*_Sox2_Sox9_overlap.bed  > ChIP-seq/processed_data/Merged_rep_Consensus5_overlap.bed
# bedtools multiinter -header -i ChIP-seq/processed_data/*_Sox2_exclusive.bed  > ChIP-seq/processed_data/Merged_rep_consensus5_Sox2_exclusive.bed
# bedtools multiinter -header -i ChIP-seq/processed_data/*_Sox9_exclusive.bed  > ChIP-seq/processed_data/Merged_rep_consensus5_Sox9_exclusive.bed

## Load consensus sets:
### Co-bound set:
co_bound_consensus <- fread("../../processed_data/Merged_rep_Consensus5_overlap.bed") %>%
  toGRanges() %>%
  reduce()
length(co_bound_consensus)

## SOX2 exclusive:
Sox2_consensus <- fread("../../processed_data/Merged_rep_consensus5_Sox2_exclusive.bed") %>%
  toGRanges() %>%
  reduce()
length(Sox2_consensus)

## SOX9 exclusive:
Sox9_consensus <- fread("../../processed_data/Merged_rep_consensus5_Sox9_exclusive.bed") %>%
  toGRanges() %>%
  reduce()
length(Sox9_consensus)

pdf("/Figures/venn_diag_co_bound.pdf", height = 3, width = 3)
draw.pairwise.venn(area1 = length(Sox2_consensus) + length(co_bound_consensus), 
                   area2 = length(Sox9_consensus) + length(co_bound_consensus), 
                   cross.area = length(co_bound_consensus), scaled =T, 
                   fill = c("darkgrey","brown"), 
                   cex = c(1.5,1.5,1.5), 
                   alpha = c(0.5,0.5), fontfamily = rep("sans",3))

dev.off()

## Circular permutation test:
### Get all SOX2 peaks:
sox2_peaks <- c(Sox2_consensus, co_bound_consensus)

### Create all SOX9 peaks:
sox9_peaks <- c(Sox9_consensus, co_bound_consensus)

### Perform overlap test:
set.seed(999)
perm_test_co_bound <- permTest(A = sox2_peaks, 
                               B = sox9_peaks,
                               ntimes = 10000,
                               randomize.function=circularRandomizeRegions, 
                               evaluate.function=numOverlaps, 
                               count.once=TRUE, 
                               genome="hg38", 
                               mc.set.seed=FALSE, 
                               mc.cores=20)
perm_test_co_bound

```

## Figure D (right): Venn diagram of consensus co-bound SOX2/SOX9 peaks vs super-enhancers in GSC:
```{r}
## Load consensus SE set: SEs occuring in >1 GSC
consensus_SE <- fread("../../processed_data/Consensus_SEs.bed") %>% 
  toGRanges() %>%
  reduce()
length(consensus_SE)

## Load consensus co-bound SOX2 and SOX9 set:
co_bound_consensus <- fread("../../processed_data/Merged_rep_Consensus5_overlap.bed") %>%
  toGRanges() %>%
  reduce()
length(co_bound_consensus)

co_bound_SOX_SEs <- subsetByOverlaps(co_bound_consensus, consensus_SE)
co_bound_SOX_not_SE <- subsetByOverlaps(co_bound_consensus, consensus_SE, invert = T)
SE_not_co_bound <- subsetByOverlaps(consensus_SE, co_bound_consensus, invert = T)

pdf("Figures/venn_diag_co_bound_SEs.pdf", height = 3, width = 3)
draw.pairwise.venn(area2 = length(co_bound_SOX_not_SE) +length(co_bound_SOX_SEs), 
                   area1 = length(SE_not_co_bound) + length(co_bound_SOX_SEs), 
                   cross.area = length(co_bound_SOX_SEs), scaled =T, 
                   cex = c(1.5,1.5,1.5), 
                   fill = c("lightblue","brown"), alpha = c(0.5,0.5), inverted = T, fontfamily = rep("sans",3))
dev.off()

### Perform overlap test:
set.seed(999)
perm_test_co_bound_SE <- permTest(A = co_bound_consensus, 
                               B = consensus_SE,
                               ntimes = 10000,
                               randomize.function=circularRandomizeRegions, 
                               evaluate.function=numOverlaps, 
                               count.once=TRUE, 
                               genome="hg38", 
                               mc.set.seed=FALSE, 
                               mc.cores=20)

```

## Figure E: GO terms for co-bound SOX2/SOX9 sites at SEs:
```{r}
### First overlap co-bound SOX2/SOX9 sets with consensus SEs and get gene annotations using GREAT via the online tool
### Then perform GO Biological term analysis using clusterProfiler.
library(clusterProfiler)
library(org.Hs.eg.db) 
library(DOSE)
 
# # Convert to ENTREZID:
# this_set_entrezid <- bitr(co_bound_SE_genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db) ## For DO terms
# nrow(this_set_entrezid)
#   
# ## GO and DO tests:
# go_bp_terms <- enrichGO(gene = this_set_entrezid$SYMBOL,
#                         OrgDb = org.Hs.eg.db,
#                         keyType = 'SYMBOL',
#                         ont = "BP",
#                         pAdjustMethod = "BH",
#                         pvalueCutoff = 0.01,
#                         qvalueCutoff = 0.05) %>% as.data.table() %>% 
#   dplyr::mutate(term = "GO Biological Process", set = "co_bound_SE_genes")
#   
# go_bp_terms %>% fwrite("../../processed_data/gene_set_enrichment_biological_processes_co-bound_sox_in_SEs.tsv", quote = F, sep = "\t", row.names = F)
go_bp_terms = fread("../../processed_data/gene_set_enrichment_biological_processes_co-bound_sox_in_SEs.tsv")

pdf("Figures/cobound_overlap_SEs_go_biological_terms.pdf", width = 4.2, height = 3.1, useDingbats = T, onefile = F)
go_bp_terms %>%
  slice_head(n=10) %>% 
  ungroup() %>%
  separate(GeneRatio, into = c("GeneCount", "GeneTotal"), sep = "/") %>%
  separate(BgRatio, into = c("BgCount", "BgTotal"), sep = "/") %>%
  dplyr::mutate(FC = (as.numeric(GeneCount)/as.numeric(GeneTotal))/(as.numeric(BgCount)/as.numeric(BgTotal))) %>%
  ggplot(aes(x = -log10(qvalue), y=reorder(Description, -qvalue), size= FC)) +
  geom_segment(aes(xend = 0, yend = Description), color = "grey", size=0.2) +
  geom_point(aes(color = set)) +
  labs(y =NULL, title = "Super-enhancers", x = NULL) +
  scale_color_manual(values = c("co_bound_SE_genes" = "#7570B3")) +
  scale_size_binned(limits=c(1,10)) +
  scale_x_reverse(expand=c(0.01,0.1), n.breaks=5, limits=c(8,0)) +
  scale_y_discrete(position = "right") +
  theme_classic() +
  coord_capped_cart(right = "none", bottom = "both") +
  theme_colplot +
  theme(strip.text = element_text(size=12, hjust=0),
        legend.position = "none",
        axis.ticks.y = element_line(linewidth = 0.2),
        axis.line.x = element_line(colour = "transparent"), 
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(), 
        plot.title = element_text(face = "plain"),
        axis.line.y = element_line(linewidth = 0.2),
        axis.text.y = element_text(size=10))
dev.off()

```

## Figure F: SOX2/SOX9 expression correlation across TCGA cancers:
```{r}
## From TCGA RNA-seq data (cbioportal):
TCGA <- read.delim("/Users/alhafidzhamdan/PhD/Data/ChIP-seq/Sox2_Sox9/External_data/cbioportal/tcga_pan_can_atlas_2018/sox_genes/cbioportal_sox2_sox9.txt", header=T)  %>%
  select(Sample = Sample.Id,
         Cancer_type = TCGA.PanCanAtlas.Cancer.Type.Acronym,
         Sox2_expression = SOX2..mRNA.Expression..RSEM..Batch.normalized.from.Illumina.HiSeq_RNASeqV2...log2.value...1..,
         Sox9_expression = SOX9..mRNA.Expression..RSEM..Batch.normalized.from.Illumina.HiSeq_RNASeqV2...log2.value...1..)

pdf("Figures/TCGA_SOX2_SOX9_exp_corr.pdf", height = 4.9, width = 6)
TCGA_stats %>% 
  ggplot(aes(x=reorder(Cancer_type,cor), fill=-log10(p.adj), y = cor)) +
  geom_col() + 
  scale_fill_viridis_c(option = "magma") + 
  theme_classic() +
  lemon::coord_capped_cart(left = "both") + 
  labs(
    # title = "SOX2 and SOX9 expression correlation analysis", 
       y="SOX2/SOX9\nexpression correlation", x = "\nTCGA cancer types", fill = expression("-log"[10]*"Q-value")) + 
  scale_y_continuous(breaks = c(-0.5,-0.25,0,0.25,0.5), limits = c(-0.5,0.5)) +
  theme(axis.text.y = element_text(size=12),
        axis.title = element_text(size=13),
        legend.position = "bottom",
        plot.title = element_text(size=13, face = "bold"),
        legend.title = element_text(size=13, face = "bold", vjust = 0.9),
        legend.text = element_text(size=11), 
        axis.text.x = element_text(size=12, angle = 90, hjust=0.9, vjust=0.5))
dev.off()

```

## Figure G: Inverted palindromic motif spacing at co-bound SOX2/SOX9 sites:
```{r}
pdf("Figures/Thesis/chapter_5/sox_monomer_pairing_inverted_palindrome_gaps.pdf", width = 2, height = 3, useDingbats = F, onefile = F)
fread("../../processed_data/SOX_monomer_at_SOX2_co_bound_ENH_spamo.tsv") %>% 
  dplyr::filter(str_detect(sec_alt, "Sox|SRY"))  %>% 
  dplyr::filter(orient == "0" | orient == "1") %>% 
  group_by(gap) %>% 
  dplyr::summarise(total = sum(count)) %>% 
  ggplot(aes(x=gap, y=total)) +
  geom_col() +
  theme_classic() +
  coord_capped_cart(left = "both", bottom = "none") +
  labs(x= "Inverted palindromic\n SOX motif gap", y = "Count") +
  theme_colplot +
  scale_y_continuous(limits=c(0,150), expand = c(0.02,0.05)) +
  scale_x_continuous(expand = c(0.02,0.01))
dev.off()

```

## Figure H: Coverage tracks of SOX2 and SOX9 near CDK6 and PTRPRZ1, USCS browser at https://genome.ucsc.edu/s/alhafidzhamdan/co_bound_SOX2_SOX9_peaks
## Figure I: Run within the MEME-Suite tool via the online platform - https://meme-suite.org/meme/tools/meme-chip (Classic mode)




